### 2019-11-14 latex/vhcs.tex
Adding more TODOs and renaming subsections
Wrote the motivation of VHCs
Also wrote a bunch more stuff for VNHCs, providing more detail for the
Hamiltonian-from-Lagrangian section.

\subsection{Review: Lagrangian Virtual Holonomic Constraints}
Take an underactuated mechanical system with Lagrangian
\begin{equation}\label{eqn:lagrangian}
\mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}^T D(q) \dot{q} - P(q)
\end{equation}
where \(q = (q_1,\ldots,q_n) \in \mathcal{Q}\) is the configuration vector, 
the inertia matrix \(D(q) = D^T(q)\) is positive definite for all 
\(q \in \mathcal{Q}\), and \(P(q)\) is the potential of the system.

The equations of motion for this system can be computed from 
the Euler-Lagrange equations (\ref{eqn:el_eqns})
\begin{equation}\label{eqn:el_eqns}
   \frac{d}{dt} \left( \frac{\partial \mathcal{L}}{\partial \dot{q}}\right) -
   \frac{\partial \mathcal{L}}{\partial q} = B(q)\tau
\end{equation}
which result in (\ref{eqn:eom_lagrangian}), where
\(C : T\mathcal{Q} \rightarrow \mathbb{R}^{n \times n}\) represents the
generalized coriolis and centrifugal forces,
\(B : \mathcal{Q} \rightarrow \mathbb{R}^{n \times m}\) is a 
\(C^1\) full rank control matrix, and \(\tau \in \mathbb{R}^{m}\) is the control 
input. \textbf{TODO: citation for E-L equations}

@@

Suppose one were to use the control input \(\tau\) to enforce a
constraint on the generalized coordinates which depends only on \(q\) and not on
\(\dot{q}\). This type of constraint is called ``holonomic"; since the 
constraint is enforced by control as opposed to being a feature of the mechanics
of the system, it is a called a ``virtual" constraint. 
This intuition leads to the following mathematical development, described in 
\cite{vhcs_for_el_systems}:

@@

\begin{defn}\label{defn:vhc_regular}
   A VHC of order \((n-1)\) is \textbf{regular} if 
   system (\ref{eqn:eom_lagrangian}) with output \(e = h(q)\) has vector
   relative degree \(\{2,2,\ldots,2\}\) on \(\Gamma\).
   Equivalently, a VHC is regular if 
   \[
      dh_q D^{-1}(q)B(q)
   \]
   is nonsingular everywhere on \(\Gamma\).
\end{defn}

@@

To show this, we repeat the derivation of
\textcite{vhcs_for_el_systems,lagrangian_structure_reduced_dynamics_vhcs} and take
derivatives of \(e\).

@@

Let \(\mu(q,\dot{q})\) be given by 
\begin{equation*}
   \mu(q,\dot{q}) := -dh_qD^{-1}(q)\left[C(q,\dot{q})\dot{q} + \nabla P(q) \right] 
      + \begin{bmatrix}
         \dot{q}^T \nabla^2 h_1(q) \dot{q} \\
         \vdots \\
         \dot{q}^T \nabla^2 h_{n-1}(q) \dot{q}
      \end{bmatrix}
\end{equation*}
Then the resulting error dynamics are
\begin{equation}\label{eqn:vhc_error_dynamics}
   \ddot{e} = \mu(q,\dot{q}) + dh_q D^{-1}B(q) \tau
\end{equation}

Since \(h\) is a regular VHC, it is possible to solve for the control input. 
The controller
\begin{equation*}
   \tau(q,\dot{q}) = \left(dh_q D^{-1}B(q)\right)^{-1}\left[-k_p e - k_d \dot{e} -
   \mu(q,\dot{q})\right]
\end{equation*}
will generate the linear error dynamics \(\ddot{e} = -k_p e - k_d\dot{e}\) 
which make \(\Gamma\) asymptotically stable.

After constructing this controller, a natural question to ask is how system
(\ref{eqn:eom_lagrangian}) behaves once it has stabilized on the constraint
manifold \(\Gamma\). For this, one must make the following assumption.

@@

As in \cite{lagrangian_structure_reduced_dynamics_vhcs}, 
now \(s \in \Theta\) be the virtual configuration which parameterizes
\(h^{-1}(0)\) through the diffeomorphism 
\begin{align*}
   \sigma : \Theta &\rightarrow h^{-1}(0) \\
   s &\mapsto q
\end{align*}
Likewise, the tangent bundle \(T\Theta\) is diffeomorphic to \(Gamma\) 
through the mapping
\begin{align*}
   d\sigma : T\Theta &\rightarrow \Gamma \\
   (s,\dot{s}) &\mapsto (q,\dot{q})
\end{align*}

Using this new virtual configuration, the state of system
(\ref{eqn:eom_lagrangian}) on the constraint manifold
is parameterized by \(q = \sigma(s)\) and \(\dot{q} = \sigma'(s)\dot{s}\).
Inserting these into (\ref{eqn:eom_lagrangian_with_B_perp}) (and dropping the
function notation on \(B\),\(D\),\(C\), and \(P\) for simplicity), one gets
\begin{align*}
   B^\perp D\left(\sigma'(s)\ddot{s} + \sigma"(s)\dot{s}^2 \right)
   + B^\perp \left[ C \sigma'(s)\dot{s} + \nabla P \right] = 0
\end{align*}

By how \(C(q,\dot{q})\) is defined, 
one can collect the product \(B^\perp C \dot{q}\) in the following way.
First, combine the Christoffel symbols  into a matrix
\(Q_i : \mathcal{Q} \rightarrow \mathbb{R}^{n \times n}\)
defined element-wise by
\begin{equation*}
   [Q_i]_{j,k}(q) = \frac{1}{2}\left(\partial_{q_k} [D]_{i,j}(q)
      + \partial_{q_j} [D]_{i,k}(q) 
      + \partial_{q_i} [D]_{k,j}(q)\right)
\end{equation*}

Then \(B^\perp C \dot{q}\) becomes
\begin{equation*}
   \sum \limits_{i=1}^n B_i^\perp \dot{q}^T Q_i
\end{equation*}

@@

>>>>

%---------- Motivation ----------%
\subsection{Motivation}
Most classic control architectures achieve one of two goals: equilibrium
stabilization, or tracking of a signal over time (which is known as
trajectory tracking). In fact, most trajectory-tracking controllers achieve
their results by changing a system's coordinates so that the new system is at
equilibrium if it is on the trajectory; so in fact, most control architectures
in the past have focused on equilibrium stabilization. \textbf{TODO: citation to
support this claim?}

While this is a perfectly valid approach for many systems, it is not sufficient
to provide robust, intricate motion on anything but the simplest of control
systems \textbf{TODO: citation? Can I suppport this?}. Take the example of an
autonomous vehicle driving on a road. If the desired path is a time trajectory 
and the robot experiences a disturbance, it will need to ``catch up" to the
trajectory it was following. This is a problem, because the vehicle will take
shortcuts that may take it off its original trajectory, which can cause
collisions with pedestrians, buildings, or simply cause the robot to speed down
the road.

\textbf{TODO figure with an example of this robot trajectory tracking issue}

For this reason, mobile roboticists use \textbf{path tracking}:
they define a constraint on the position of the vehicle, which does not depend on
time. Rather than catching up to where the robot is ``supposed to be" in time, a
path-tracking controller drives the robot to the nearest point on the path. This
is a much more robust control architecture, as it forces the robot to stay
within the confines of the path, so that it does not accidentally kill an
unsuspecting pedestrian.

\textbf{TODO: figure showing path tracking}

Take now the example of a walking robot. If the legs are given a trajectory in
time to follow, then the second the robot stumbles or slips it will try to
``catch up" to the trajectory and the robot will fall over \textbf{TODO: cite
grizzle. Does he make this argument too?}.  To fix this issue, one requires a
path-tracking controller for general robotic systems. An engineer should be able
to provide a constraint on the robot's configuration which does not depend on
time, and the controller should keep the robot on that constraint. This is
precisely what Virtual Holonomic Constraints attempt to accomplish.

%---------- Define VHCs ----------%
\subsection{Lagrangian Virtual Holonomic Constraints}
Take an underactuated mechanical system with Lagrangian
\begin{equation}\label{eqn:lagrangian}
\mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}^T D(q) \dot{q} - P(q)
\end{equation}
where \(q = (q_1,\ldots,q_n) \in \mathcal{Q}\) is the configuration vector, 
the inertia matrix \(D(q) = D^T(q) \mathbb{R}^{n \times n}\) 
is positive definite for all 
\(q \in \mathcal{Q}\), and \(P(q) \in \mathbb{R}\) is the potential of the
system.

The Euler-Lagrange equations 
\begin{equation}\label{eqn:el_eqns}
   \frac{d}{dt} \left( \frac{\partial \mathcal{L}}{\partial \dot{q}}\right) -
   \frac{\partial \mathcal{L}}{\partial q} = B(q)\tau
\end{equation}
produce the equations of motion (\ref{eqn:eom_lagrangian}), where
\(C : T\mathcal{Q} \rightarrow \mathbb{R}^{n \times n}\) represents the
generalized coriolis and centrifugal forces,
\(B : \mathcal{Q} \rightarrow \mathbb{R}^{n \times m}\) is a 
\(C^1\) full rank control matrix, and \(\tau \in \mathbb{R}^{m}\) is the control 
input. \textbf{TODO: citation for E-L equations}

@@

Suppose one were to use the control input \(\tau\) to enforce a
constraint on \(q\), which does not depend on \(\dot{q}\). This type of
constraint is called ``holonomic". Furthermore, since the constraint is enforced
by control as opposed to being a feature of the mechanics of the system, it is a
called a ``virtual" constraint.  This intuition leads to the following
mathematical development, as was described in \cite{vhcs_for_el_systems}:

@@

\begin{defn}\label{defn:vhc_regular}
   A VHC \(h(q)\) of order \((n-1)\) is \textbf{regular} if 
   system (\ref{eqn:eom_lagrangian}) with output \(e = h(q)\) has vector
   relative degree \(\{2,2,\ldots,2\}\) on \(\Gamma\).
   Equivalently, a VHC is regular if and only if 
   \[
      dh_q D^{-1}(q)B(q)
   \]
   is full rank everywhere on \(\Gamma\).
\end{defn}

@@

To show this, we repeat the analysis by
\textcite{vhcs_for_el_systems,lagrangian_structure_reduced_dynamics_vhcs} and take
derivatives of \(e\).

@@

Let \(\mu(q,\dot{q})\) be given by 
\begin{equation*}
   \mu(q,\dot{q}) := -dh_qD^{-1}(q)\left[C(q,\dot{q})\dot{q} + \nabla P(q) \right] 
      + \begin{bmatrix}
         \dot{q}^T \nabla^2 h_1(q) \dot{q} \\
         \vdots \\
         \dot{q}^T \nabla^2 h_{n-1}(q) \dot{q}
      \end{bmatrix}
\end{equation*}
Then the resulting output dynamics are
\begin{equation}\label{eqn:vhc_error_dynamics}
   \ddot{e} = \mu(q,\dot{q}) + dh_q D^{-1}B(q) \tau
\end{equation}

Since \(h\) is a regular VHC, it is possible to solve for the control input. 
The controller
\begin{equation*}
   \tau(q,\dot{q}) = \left(dh_q D^{-1}B(q)\right)^{-1}\left[-k_p e - k_d \dot{e} -
   \mu(q,\dot{q})\right]
\end{equation*}
will generate the linear output dynamics \(\ddot{e} = -k_p e - k_d\dot{e}\) 
which make \(\Gamma\) asymptotically stable.

%---------- Constrained Dynamics ----------%
\subsection{Constrained Dynamics for Lagrangian VHCs}
After constructing this controller, a natural question to ask is how system
(\ref{eqn:eom_lagrangian}) behaves once it has stabilized on the constraint
manifold \(\Gamma\). For this, one must make the following assumption.

@@

As in \cite{lagrangian_structure_reduced_dynamics_vhcs}, 
let \(s \in \Theta\) be the virtual configuration which parameterizes
\(h^{-1}(0)\) through a diffeomorphism 
\begin{align*}
   \sigma : \Theta &\rightarrow h^{-1}(0) \\
   s &\mapsto \sigma(s) = q
\end{align*}
Likewise, the tangent bundle \(T\Theta\) is diffeomorphic to \(\Gamma\) 
through the mapping
\begin{align*}
   d\sigma : T\Theta &\rightarrow \Gamma \\
   (s,\dot{s}) &\mapsto (q,\dot{q})
\end{align*}

Using this new virtual configuration, the state of system
(\ref{eqn:eom_lagrangian}) on \(\Gamma\) 
is parameterized by \(q = \sigma(s)\) and \(\dot{q} = \sigma'(s)\dot{s}\).
Inserting these into (\ref{eqn:eom_lagrangian_with_B_perp}) (and dropping the
function notation on \(B\),\(D\),\(C\), and \(P\) for simplicity), one gets
\begin{align*}
   B^\perp D\left(\sigma'(s)\ddot{s} + \sigma"(s)\dot{s}^2 \right)
   + B^\perp \left[ C \sigma'(s)\dot{s} + \nabla P \right] = 0
\end{align*}

The matrix \(C(q,\dot{q})\) is defined element-wise by 
\begin{align*}
   [C]_{j,k} &= \sum \limits_{i=1}^n c_{i,j,k}\dot{q}_i \\
   c_{i,j,k} &= ???
\end{align*}
where \(c_{i,j,k}\) are the Christoffel symbols.
As such, one can collect the product \(B^\perp C \dot{q}\) in the following way.
First, combine the Christoffel symbols  into matrices
\(Q_i : \mathcal{Q} \rightarrow \mathbb{R}^{n \times n}\)
defined element-wise by
\begin{equation*}
   [Q_i]_{j,k}(q) = \frac{1}{2}\left(\partial_{q_k} [D]_{i,j}(q)
      + \partial_{q_j} [D]_{i,k}(q) 
      + \partial_{q_i} [D]_{k,j}(q)\right)
\end{equation*}
Then \(B^\perp C \dot{q}\) becomes
\begin{equation*}
   \sum \limits_{i=1}^n B_i^\perp \dot{q}^T Q_i
\end{equation*}

@@

%---------- Pros and Cons of VHCs ----------%
\subsection{Benefits and Drawbacks of VHCs}
\textbf{TODO: why are VHCs good? Who has used them? What is their downside?}

### 2019-11-14 latex/vnhcs.tex
Adding more TODOs and renaming subsections
Wrote the motivation of VHCs
Also wrote a bunch more stuff for VNHCs, providing more detail for the
Hamiltonian-from-Lagrangian section.

%/----- Motivation -----/%
\subsection{Motivation}
Suppose one wishes to set a constraint that depends on both the configuration
and the velocity.
TODO: Why do we bother with Hamiltonian? Why can't we do virtual nonholonomic 
constraints in Lagrangian?

\subsection{Review: Hamiltonian Systems}
\textbf{TODO: What is a Hamiltonian system and why does it matter?}

One can compute the Hamiltonian of the Lagrangian (\ref{eqn:lagrangian}) by 
performing a Legendre transform on the Lagrangian. 
First, define the conjugate of momenta for \(q\) by 
\begin{equation*}
p = \frac{\partial \mathcal{L}}{\partial \dot{q}} \in \mathbb{R}^n
\end{equation*}
Then, the Legendre transform is performed by taking
\begin{equation*}
\mathcal{H}(q,p) = p^T \dot{q} - \mathcal{L}(q,\dot{q})
\end{equation*}
It is easy to verify that the Hamiltonian for system (\ref{eqn:lagrangian}) is 
given by the total mechanical energy (\ref{eqn:hamiltonian}).
\begin{equation}\label{eqn:hamiltonian}
\mathcal{H}(q,p) = E(q,p) = \frac{1}{2} p^T M^{-1}(q) p + V(q)
\end{equation}

The equations of motion for the system in Hamiltonian coordinates is given by
\begin{align}\label{eqn:hamiltionian_eom}
\begin{split}
\dot{q} &= \frac{\partial \mathcal{H}}{\partial p} = M^{-1}(q) p \\
\dot{p} &= -\frac{\partial \mathcal{H}}{\partial q} + B \tau
\end{split}
\end{align}

@@

\subsection{Hamiltonian Virtual Nonholonomic Constraints}
While VHCs are still possible in the Hamiltonian framework, the assumptions
required to make this work are slightly different. To see this, let us look back
to the derivation which led to the error dynamics (\ref{eqn:vhc_error_dynamics})
and attempt to reconstruct this in Hamiltonian coordinates.

Suppose the mechanical system has degree of underactuation one, so that coordinates of the system can be split into an unactuated component \(q_u \in [\mathbb{R}]_T, \, T \in \mathbb{R}_{>0}\) which is not influence by control, along with an actuated component \(q_a\); that is, suppose \(B\) is of the form \(B(q) = [0_m, B_1^T(q), \ldots, B_n^T(q)]^T, \, B_i^T(q) \in \mathbb{R}^{n-1}\) and \(\tau \in \mathbb{R}^{n-1}\). In this case, \(q = (q_u, q_a)^T\) and the equations of motion become
\begin{align}\label{eqn:unactuated_actuated_eom}
\begin{split}
\dot{q_u} &= e_1^T M^{-1}(q) p \\
\dot{p_u} &= -p^T\frac{\partial M}{\partial q_u} p - \partial_{q_u}V(q) \\
\dot{q_a} &= 
\begin{bmatrix}
0 \cdots 0 \\
I_{n-1} \\
\end{bmatrix} M^{-1}(q) p \\
\dot{p_a} &= -p^T\frac{\partial M}{\partial q_a} p - \nabla{q_a}V(q) + 
\begin{bmatrix}
B_1(q) \\
\vdots \\
B_n(q)
\end{bmatrix} \tau
\end{split}
\end{align}

@@

Now we solve for \(\tau\) by finding \(\ddot{e}\).
\begin{align*}
    e &= h(q,p) = qa - f(q_u,p_u)\\
    \Rightarrow \dot{e} &= \dot{q_a} - df_{q_u}\dot{q_u} -df_{p_u}\dot{p_u} \\
    &= [-df_{q_u} I_{n-1}]\dot{q} - df_{p_u} \dot{p_u} \\
    &= dh_q M^{-1}(q) p - df_{p_u}\left( -\frac{1}{2}p^T \frac{\partial M^{-1}(q)}{\partial q_u} p - \partial_{q_u}V(q) \right) \\
    &= dh_q M^{-1}(q) p + \frac{1}{2}df_{p_u} p^T \frac{\partial M^{-1}(q)}{\partial q_u} p + df_{p_u}\partial_{q_u}V(q)
\end{align*}
The control input \(\tau\) only appears in \(\dot{p_a}\) (see (\ref{eqn:unactuated_actuated_eom})). To simplify the analysis, terms in \(\ddot{e}\) which do not depend on \(\dot{p_a}\) explicitly are lumped together under the symbol \((*)\):
\begin{align*}
    \ddot{e} &= dh_q M^{-1}(q) \dot{p} + df_{p_u} p^T \frac{\partial M^{-1}(q)}{\partial q_u} \dot{p} + (*) \\
    &= (dh_q M^{-1}(q) + df_{p_u} p^T \frac{\partial M^{-1}(q)}{\partial q_u})B\tau + (*) \\
    &= (dh_q M^{-1}(q) + dh_{p_u} p^T \frac{\partial M^{-1}(q)}{\partial q_u})B\tau + (*)
\end{align*}

From the derivations above, one can solve for \(\tau\) iff the matrix on the left of \(\tau\) is full rank. Thus, for systems with degree of underactuation one we give the following definition.
\begin{defn}
A VNHC \(h(q,p) = 0\) of order \(n - 1\) is \textbf{regular} if \(dh_{p_a} = 0\), \(dh_{q_a} = (1 \ldots 1)^T\), and 
\[
\text{rank}\left\{ (dh_q M^{-1}(q) + dh_{p_u} p^T \frac{\partial M^{-1}(q)}{\partial q_u})B\right\} = n - 1
\]
everywhere on the constraint manifold \(\Gamma\). Equivalently, a VNHC \(h\) of order \(n - 1\) is regular if it satisfies Assumption \ref{assm:vnhc_is_on_qu_pu} and system (\ref{eqn:unactuated_actuated_eom}) with output \(e = h(q,p)\) is of relative degree \(\{2,2,\ldots,2\}\) everywhere on \(\Gamma\).
\end{defn}

In general, \(\dot{e}\) is a function of \(q_u\) and \(p = (p_u,p_a)^T\). Since the purpose of a regular VNHC is to fully parameterize \(\Gamma\) by \((q_u,p_u)\), it is essential that one can solve for \(p_a = p_a(q_u,p_u)\). Unfortunately this often cannot be done, since \(\dot{e}\) contains the quadratic term
\[
\frac{1}{2} df_{p_u} p^T \frac{\partial M^{-1}(q)}{\partial q_u} p 
\]
We can solve for \(p_a\) if this quadratic term does not exist.
\begin{assm}\label{assm:M_is_Mqa}
Assume \(\partial M(q) / \partial q_u = 0 \Leftrightarrow \partial M^{-1}(q) / \partial q_u = 0\)
\end{assm}

Under Assumption \ref{assm:M_is_Mqa}, we get that the rank condition for \(h(q,p)\) to be a regular VNHC reduces to \(\text{rank}\left(dh_q M^{-1} B\right) = n - 1\). This is the same rank condition as required for Virtual Holonomic Constraints.

Now we solve for \(p_a\) on the constraint manifold (when \(e = \dot{e} = 0\)):
\begin{align*}
    \dot{e} = dh_q M^{-1}(q)p + df_{p_u} \partial_{q_u}V(q) &= 0\\
    \Leftrightarrow dh_q M^{-1}(q)e_1 p_u + dh_q M^{-1}(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix} p_a &= -df_{p_u} \partial_{q_u}V(q) \\
\end{align*}
\begin{align*}
    \Leftrightarrow dh_q M^{-1}(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix} p_a = -\left(df_{p_u}\partial_{q_u}V(q) + dh_q M^{-1}(q)e_1 p_u\right)
\end{align*}
One can linearly solve for \(p_a\) if and only if the matrix in front of it is invertible.

This leads us to a natural definition.
\begin{defn}\label{defn:solvability}
A VNHC \(h(q,p)\) is \textbf{solvable} (NOTE: actionable? what's a good name?) if
\[
\text{rank}\left(dh_q M^{-1}(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix}\right) = n - 1
\]
\end{defn}

@@

\begin{thm}\label{thm:equation_for_pa}
Suppose assumptions \ref{assm:vnhc_is_on_qu_pu} and \ref{assm:M_is_Mqa} hold.
If a regular VNHC \(h(q,p) = q_a - f(q_u,p_u)\) is solvable, then the
parameterization for \(p_a\) on the constraint manifold is given by
\begin{align*}
p_a &= -\left(dh_q M^{-1}(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix}\right)^{-1}\left( df_{p_u}\partial_{q_u}V(q) + dh_q M^{-1}(q)e_1 p_u\right) \\
    &=: g(q_u,p_u)
\end{align*}
and the constrained dynamics on \(\Gamma\) are given by
\begin{align*}
    \dot{q_u} &= e_1^T M^{-1}(q_a) \begin{bmatrix}
    p_u \\
    p_a
    \end{bmatrix}\mid_{q_a = f(q_u,p_u), p_a = g(q_u,p_u)} \\
    \dot{p_u} &= -\partial_{q_u} V(q_u,q_a) \mid_{q_a = f(q_u,p_u)}
\end{align*}
\end{thm}

@@

%---------- Hamiltonian VHCS ---------%
\subsection{Hamiltonian Virtual Holonomic Constraints}
\textbf{TODO: Why can we not do the standard VHC approach for Hamiltonian? 
Show how to use the above to make it work}

>>>>

%----- Motivation -----%
\subsection{Motivation}
\textbf{TODO: Why do we bother with Hamiltonian? Why can't we do virtual nonholonomic 
constraints in Lagrangian? What are some use-cases where VHCs don't work?}

%---------- Hamiltonian from Lagrangian ----------%
\subsection{Review: Hamiltonian Systems}
\textbf{TODO: What is a Hamiltonian system and why does it matter?}

One can compute the Hamiltonian of a system by performing a Legendre transform
on its Lagrangian \textbf{TODO: citation for legendre transform}.
First, define the conjugate of momenta for \(q\) by 
\begin{equation*}
p = \frac{\partial \mathcal{L}}{\partial \dot{q}} \in \mathbb{R}^n
\end{equation*}
Then, the Legendre transform is performed by taking
\begin{equation*}
\mathcal{H}(q,p) = p^T \dot{q} - \mathcal{L}(q,\dot{q})
\end{equation*}

For a mechanical system (\ref{eqn:lagrangian}), the conjugate of momenta for
\(q\) is given by 
\begin{equation*}
    p = M(q)\dot{q}
\end{equation*}
which means the Hamiltonian of the system is given by
the total mechanical energy \(E\) (\ref{eqn:hamiltonian} in 
\((q,p)\) coordinates.
\begin{equation}\label{eqn:hamiltonian}
\mathcal{H}(q,p) = E(q,p) = \frac{1}{2} p^T \Minv(q) p + V(q)
\end{equation}
Note that \(M(q)\) is the inertia matrix and \(V(q)\) is the potential for
the Hamiltonian system. This is simply a trick to distinguish them notationally
from the Lagrangian \(D(q)\) and \(P(q)\); they are, in fact, identical in their
contents.

The equations of motion for the system in Hamiltonian coordinates is given by
\begin{align}\label{eqn:hamiltionian_eom}
\begin{split}
\dot{q} &= \frac{\partial \mathcal{H}}{\partial p} = \Minv(q) p \\
\dot{p} &= -\frac{\partial \mathcal{H}}{\partial q} + B \tau
\end{split}
\end{align}

@@

%---------- Hamiltonian VNHCs ----------%
\subsection{Hamiltonian Virtual Nonholonomic Constraints}
While VHCs are still possible in the Hamiltonian framework, the assumptions
required to make this work are slightly different. Rather than deriving
Hamiltonian VHCs directly, we will produce results for nonholonomic constraints
first as VHCs are a special case of this new framework.

Suppose the mechanical system has degree of underactuation one, so that coordinates of the system can be split into an unactuated component \(q_u \in [\mathbb{R}]_T, \, T \in \mathbb{R}_{>0}\) which is not influence by control, along with an actuated component \(q_a\); that is, suppose \(B\) is of the form \(B(q) = [0_m, B_1^T(q), \ldots, B_n^T(q)]^T, \, B_i^T(q) \in \mathbb{R}^{n-1}\) and \(\tau \in \mathbb{R}^{n-1}\). In this case, \(q = (q_u, q_a)^T\) and the equations of motion become
\begin{align}\label{eqn:unactuated_actuated_eom}
\begin{split}
\dot{q_u} &= e_1^T \Minv(q) p \\
\dot{p_u} &= -p^T\frac{\partial M}{\partial q_u} p - \partial_{q_u}V(q) \\
\dot{q_a} &= 
\begin{bmatrix}
0 \cdots 0 \\
I_{n-1} \\
\end{bmatrix} \Minv(q) p \\
\dot{p_a} &= -p^T\frac{\partial M}{\partial q_a} p - \nabla{q_a}V(q) + 
\begin{bmatrix}
B_1(q) \\
\vdots \\
B_n(q)
\end{bmatrix} \tau
\end{split}
\end{align}

@@

Now we solve for \(\tau\) by finding \(\ddot{e}\).
\begin{align*}
    e &= h(q,p) = qa - f(q_u,p_u)\\
    \Rightarrow \dot{e} &= \dot{q_a} - df_{q_u}\dot{q_u} -df_{p_u}\dot{p_u} \\
    &= [-df_{q_u} I_{n-1}]\dot{q} - df_{p_u} \dot{p_u} \\
    &= dh_q \Minv(q) p - df_{p_u}\left( -\frac{1}{2}p^T \frac{\partial \Minv(q)}{\partial q_u} p - \partial_{q_u}V(q) \right) \\
    &= dh_q \Minv(q) p + \frac{1}{2}df_{p_u} p^T \frac{\partial \Minv(q)}{\partial q_u} p + df_{p_u}\partial_{q_u}V(q)
\end{align*}
The control input \(\tau\) only appears in \(\dot{p_a}\) (see (\ref{eqn:unactuated_actuated_eom})). To simplify the analysis, terms in \(\ddot{e}\) which do not depend on \(\dot{p_a}\) explicitly are lumped together under the symbol \((*)\):
\begin{align*}
    \ddot{e} &= dh_q \Minv(q) \dot{p} + df_{p_u} p^T \frac{\partial \Minv(q)}{\partial q_u} \dot{p} + (*) \\
    &= (dh_q \Minv(q) + df_{p_u} p^T \frac{\partial \Minv(q)}{\partial q_u})B\tau + (*) \\
    &= (dh_q \Minv(q) + dh_{p_u} p^T \frac{\partial \Minv(q)}{\partial q_u})B\tau + (*)
\end{align*}

From the derivations above, one can solve for \(\tau\) iff the matrix on the left of \(\tau\) is full rank. Thus, for systems with degree of underactuation one we give the following definition.
\begin{defn}
A VNHC \(h(q,p) = 0\) of order \(n - 1\) is \textbf{regular} if \(dh_{p_a} = 0\), \(dh_{q_a} = (1 \ldots 1)^T\), and 
\[
\text{rank}\left\{ (dh_q \Minv(q) + dh_{p_u} p^T \frac{\partial \Minv(q)}{\partial q_u})B\right\} = n - 1
\]
everywhere on the constraint manifold \(\Gamma\). Equivalently, a VNHC \(h\) of order \(n - 1\) is regular if it satisfies Assumption \ref{assm:vnhc_is_on_qu_pu} and system (\ref{eqn:unactuated_actuated_eom}) with output \(e = h(q,p)\) is of relative degree \(\{2,2,\ldots,2\}\) everywhere on \(\Gamma\).
\end{defn}

In general, \(\dot{e}\) is a function of \(q_u\) and \(p = (p_u,p_a)^T\). Since the purpose of a regular VNHC is to fully parameterize \(\Gamma\) by \((q_u,p_u)\), it is essential that one can solve for \(p_a = p_a(q_u,p_u)\). Unfortunately this often cannot be done, since \(\dot{e}\) contains the quadratic term
\[
\frac{1}{2} df_{p_u} p^T \frac{\partial \Minv(q)}{\partial q_u} p 
\]
We can solve for \(p_a\) if this quadratic term does not exist.
\begin{assm}\label{assm:M_is_Mqa}
Assume \(\partial M(q) / \partial q_u = 0 \Leftrightarrow \partial \Minv(q) / \partial q_u = 0\)
\end{assm}

Under Assumption \ref{assm:M_is_Mqa}, we get that the rank condition for \(h(q,p)\) to be a regular VNHC reduces to \(\text{rank}\left(dh_q \Minv B\right) = n - 1\). This is the same rank condition as required for Virtual Holonomic Constraints.

Now we solve for \(p_a\) on the constraint manifold (when \(e = \dot{e} = 0\)):
\begin{align*}
    \dot{e} = dh_q \Minv(q)p + df_{p_u} \partial_{q_u}V(q) &= 0\\
    \Leftrightarrow dh_q \Minv(q)e_1 p_u + dh_q \Minv(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix} p_a &= -df_{p_u} \partial_{q_u}V(q) \\
\end{align*}
\begin{align*}
    \Leftrightarrow dh_q \Minv(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix} p_a = -\left(df_{p_u}\partial_{q_u}V(q) + dh_q \Minv(q)e_1 p_u\right)
\end{align*}
One can linearly solve for \(p_a\) if and only if the matrix in front of it is invertible.

This leads us to a natural definition.
\begin{defn}\label{defn:solvability}
A VNHC \(h(q,p)\) is \textbf{solvable} (NOTE: actionable? what's a good name?) if
\[
\text{rank}\left(dh_q \Minv(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix}\right) = n - 1
\]
\end{defn}

@@

\begin{thm}\label{thm:equation_for_pa}
Suppose assumptions \ref{assm:vnhc_is_on_qu_pu} and \ref{assm:M_is_Mqa} hold.
If a regular VNHC \(h(q,p) = q_a - f(q_u,p_u)\) is solvable, then the
parameterization for \(p_a\) on the constraint manifold is given by
\begin{align*}
p_a &= -\left(dh_q \Minv(q) \begin{bmatrix}
    0 & \cdots & 0 \\
    & I_{n-1} & \\
    \end{bmatrix}\right)^{-1}\left( df_{p_u}\partial_{q_u}V(q) + dh_q \Minv(q)e_1 p_u\right) \\
    &=: g(q_u,p_u)
\end{align*}
and the constrained dynamics on \(\Gamma\) are given by
\begin{align*}
    \dot{q_u} &= e_1^T \Minv(q_a) \begin{bmatrix}
    p_u \\
    p_a
    \end{bmatrix}\mid_{q_a = f(q_u,p_u), p_a = g(q_u,p_u)} \\
    \dot{p_u} &= -\partial_{q_u} V(q_u,q_a) \mid_{q_a = f(q_u,p_u)}
\end{align*}
\end{thm}

@@

%---------- Hamiltonian VHCS ---------%
\subsection{Restriction to Hamiltonian VHCs}
\textbf{TODO: Why can we not do the standard VHC approach for Hamiltonian? 
Show how to use the above to make it work}
